---
title: "Intro to stars package geospatial analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)
```

For this project we will be investigating the surface-level aerosol data that NASA makes publicly available. This time we will focus on 2019. Aerosol data is from NASA, originally downloaded from the following source: https://sedac.ciesin.columbia.edu/data/set/sdei-global-annual-gwr-pm2-5-modis-misr-seawifs-aod-v4-gl-03/data-download

From the data set description: "The PM2.5 grids consist of concentrations (micrograms per cubic meter) of all composition ground-level fine particulate matter (PM2.5) per 0.01 degree grid cells for each of the twenty-two years for the period 1998 to 2019."

We will also be making use of US census data for a shape file that provides the border of the United States. The files can be found here: https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html

Another method for obtaining US census shape files will be discussed later, in which we make use of the Tigris package.

```{r, message = FALSE}
library(stars)
library(tidyverse)
library(ggpubr)
library(terra)
library(tigris)


aero <- read_stars('sdei-global-annual-gwr-pm2-5-modis-misr-seawifs-aod-v4-gl-03-2019.tif')
shp <- st_read("cb_2018_us_nation_5m.shp")

#this is taking the shapefile for all US territories, and cutting it down to the eastern seaboard.
eastern_shp <- st_crop(shp, c(xmin=-83, xmax=-60, ymin=24, ymax=50))

plot(eastern_shp$geometry)
```



```{r, eval = F, echo = T}
#if we attempt to crop the raster file to our shapefile, we get the following error: "for cropping, the CRS of both objects has to be identical". To solve this, we need to set them to the same CRS (coordinate reference system). These are just different ways of doing map projections, and fixing this is simple.

st_crs(aero)
#the raster file is in the "WGS 84" CRS - which is the "World Geodetic System 1984"

st_crs(eastern_shp)
#the shape file is in the "NAD83" CRS - which is "North American Datum 1983"
```


```{r}
#we will transform the shape file to have the same CRS as the 
eastern_shp <- st_transform(eastern_shp, crs = st_crs(aero))

#now for comparison, we can look at the difference between the two shape files (if any)
old_shp <- st_crop(shp, c(xmin=-83, xmax=-60, ymin=24, ymax=50))

old <- ggplot()+
  geom_sf(aes(geometry = eastern_shp$geometry))+
  theme_bw()
  
new <- ggplot()+
  geom_sf(aes(geometry = old_shp$geometry))+
  theme_bw()

ggarrange(old, new)
#no apparent difference between the two plots, as these two projections must have treated this region in roughly the same way. 
```



Now to crop our global aerosol data just to the region that we are interested in. 
```{r}
eastern_aerosols <- st_crop(aero, eastern_shp)
```

Plotting fine resolution raster data can be time consuming, and downsampling is a way that we can mitigate that at the cost of resolution. Next we will plot our data using different downsampling options. 

```{r}
#leaving off the first two options as they take a very long time to run, however the code is here if you would like to see higher resolution options.

# start <- Sys.time()
# d0 <- ggplot()+
#   geom_stars(data = eastern_aerosols)+
#   geom_sf(data = eastern_shp, alpha = 0)+
#   scale_fill_viridis_c(option = "turbo", na.value = "transparent")+
#   labs(fill = "Air Particle Conc (ug/m3)")+
#   theme_void()+
#   theme(legend.position = "none")
# end <- Sys.time()
# ds0_time <- end - start
# 
# 
# start <- Sys.time()
# d1 <- ggplot()+
#   geom_stars(data = eastern_aerosols, downsample = 1)+
#   geom_sf(data = eastern_shp, alpha = 0)+
#   scale_fill_viridis_c(option = "turbo", na.value = "transparent")+
#   labs(fill = "Air Particle Conc (ug/m3)")+
#   theme_void()+
#   theme(legend.position = "none")
# end <- Sys.time()
# ds1_time <- end - start


start <- Sys.time()
d3 <- ggplot()+
  geom_stars(data = eastern_aerosols, downsample = 3)+
  geom_sf(data = eastern_shp, alpha = 0)+
  scale_fill_viridis_c(option = "turbo", na.value = "transparent")+
  labs(fill = "Air Particle Conc (ug/m3)")+
  theme_void()+
  theme(legend.position = "none")
end <- Sys.time()
ds3_time <- end - start


start <- Sys.time()
d7 <- ggplot()+
  geom_stars(data = eastern_aerosols, downsample = 7)+
  geom_sf(data = eastern_shp, alpha = 0)+
  scale_fill_viridis_c(option = "turbo", na.value = "transparent")+
  labs(fill = "Air Particle Conc (ug/m3)")+
  theme_void()+
  theme(legend.position = "none")
end <- Sys.time()
ds7_time <- end - start


start <- Sys.time()
d15 <- ggplot()+
  geom_stars(data = eastern_aerosols, downsample = 15)+
  geom_sf(data = eastern_shp, alpha = 0)+
  scale_fill_viridis_c(option = "turbo", na.value = "transparent")+
  labs(fill = "Air Particle Conc (ug/m3)")+
  theme_void()+
  theme(legend.position = 'none')
end <- Sys.time()
ds15_time <- end - start

#we could arrange these plots using ggarrange or plot.grid, however, for ease of seeing the full resolution I will plot them separately.
# ggarrange(d0, d1, d3, d7, d15, nrow = 1,  labels = c(round(ds0_time,2), round(ds1_time,2), round(ds3_time,2), round(ds7_time,2), round(ds15_time,2)))

# d0 + labs(title = paste("Time:",round(ds0_time, 2),"Seconds"))
# d1 + labs(title = paste("Time:",round(ds1_time, 2),"Seconds"))
d3 + labs(title = paste("Time:",round(ds3_time, 2),"Seconds"))
d7 + labs(title = paste("Time:",round(ds7_time, 2),"Seconds"))
d15 + labs(title = paste("Time:",round(ds15_time, 2),"Seconds"))
```

The visual change in resolution is most apparent between 7 and 15. We will use downsampling in the following section to vastly reduce the time it takes to display larger maps.

From this image we can see that the region of the eastern United States with the greatest aerosol pollution in 2019 was greatest around southeastern Pennsylvania and New Jersey, as well as northern Ohio. 

Now let's compare the east coast to the entire USA. First we will crop our vector shapefile to contain just the United States, and then we will use that to crop our stars (raster) data to that region. 

```{r}
us_shp <- st_crop(shp, c(xmin=-125, xmax=-60, ymin=22, ymax=50))

#Plotting our shape file to ensure that it covers the entire continental US
plot(us_shp$geometry)

us_shp <- st_transform(us_shp$geometry, crs = st_crs(aero))

us_aerosols <- st_crop(aero, us_shp)
```



```{r}
doParallel::registerDoParallel()

ggplot()+
  geom_stars(data = us_aerosols, downsample = 14)+
  geom_sf(data = us_shp, alpha = 0)+
  scale_fill_viridis_c(option = "turbo", na.value = "transparent")+
  labs(fill = "Air Particle Conc (ug/m3)")+
  theme_void()
```

We can see that the majority of aerosols over the United States are concentrated within the middle of the country around Ohio, with a few hotspots in the southwest and northeast. Now that we have identified the areas with the greatest air pollution in 2019, it would be interesting to determine what was driving that pollution. 

The tigris library downloads shape files for many different US census locations. Let's take a closer look at the state of Michigan, aggregating our aerosol data by county to identify the regions that are the most polluted.

```{r}
mi <- counties(state = "MI", cb = T, resolution = "20m") %>%
  st_transform(st_crs(aero))

mi_agg <- aggregate(aero, by = mi$geometry, mean, na.rm = T)

mi %>%
  mutate(aerosols = mi_agg$`sdei-global-annual-gwr-pm2-5-modis-misr-seawifs-aod-v4-gl-03-2019.tif`) %>%
  ggplot()+
  geom_sf(aes(fill = aerosols))+
  scale_fill_viridis_c(option = 'turbo')+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  labs(fill = "Aerosol Conc (ug/m3)", title = "Michigan 2019 Air Pollution")
```


From this analysis it is clear that the pollution is concentrated in the southeastern corner of the state, dissipating as you move north and west. Wayne County - the county where Detroit is located - is the area with the most pollution. And the air pollution from the city appears to extend out for several counties.

Another question we might have is which states suffer from the greatest levels of pollution. To investigate that, we can perform the same aggregation, this time looking at each of the states.

```{r}
us_state <- states(cb = T, resolution = "20m") %>%
  st_crop(c(xmin=-125, xmax=-60, ymin=22, ymax=50)) %>%
  st_transform(st_crs(aero))

#Now I will be downsampling our stars object to a lower resolution, which makes it a lot faster to aggregate over.
aero_ds <- st_downsample(aero, 15)

us_agg <- aggregate(aero_ds, by = us_state$geometry, mean, na.rm = T)

us_state %>%
  mutate(aerosols = us_agg$`sdei-global-annual-gwr-pm2-5-modis-misr-seawifs-aod-v4-gl-03-2019.tif`) %>%
  ggplot()+
  geom_sf(aes(fill = aerosols))+
  scale_fill_viridis_c(option = 'turbo')+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  labs(fill = "Aerosol Conc (ug/m3)", title = "US 2019 Air Pollution")
```

From here we can see that the most polluted state on average is Illinois. Let's do one more investigation on a state level, looking at several states now instead of just one. This region is called the "East North Central Division".

Again we can use the Tigris package.

```{r}
polluted_st <- counties(state = c("OH", "MI", "IL", "IN", "WI"), cb = T, resolution = "5m") %>%
  st_transform(st_crs(aero_ds))

polluted_agg <- aggregate(aero_ds, polluted_st, mean, na.rm = T)

polluted_st %>%
  mutate(aerosols = polluted_agg$`sdei-global-annual-gwr-pm2-5-modis-misr-seawifs-aod-v4-gl-03-2019.tif`) %>%
  ggplot()+
  geom_sf(aes(fill = aerosols))+
  scale_fill_viridis_c(option = 'turbo')+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  labs(fill = "Aerosol Conc (ug/m3)", title = "East North Central Division 2019 Air Pollution")

```


There are generally hotspots around city centers as we might expect, with more pollution near Chicago and Detroit. There is one exception in the southern portion of Indiana, which is not a metropolitan area. 

Now that we have covered basic visualizations, downsampling, and aggregation, in the next file we will move forward into interpolation and modeling. 


